---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center",
  dpi = 300
)
```

# Multivariate Bias Correction using GAMs and Nested Vine Copulas (NVC)

<!-- badges: start -->
<!-- badges: end -->

GN-VBC is a multivariate bias correction method for climate projections 
explicitly designed to preserve inter-variable, spatial and temporal dependence structures. The method is based on VBC by Funk et al. (2025) integrating Probability Integral Transforms (PIT) derived from Generalized additive models (GAMs), and nested vine copulas (NVC), a novel vine merging technique integrating two layers of dependence: spatial and inter-variable. It is designed for potentially zero-inflated climate variables, such as precipitation, and can be applied to multiple variables and locations simultaneously.

## Installation

You can install the development version of GNVBC from [GitHub](https://github.com/) with:

``` r
remotes::install_github("TheresaMeier/GNVBC")
```

## Example

Within the package, we provide an example dataset consisting of gridded reference and model data for 22 stations in the canton of Vaud, Switzerland, and four variables: temperature, relative humidity, wind speed and surface pressure. We use [Cordex](https://cds.climate.copernicus.eu/datasets/projections-cordex-domains-single-levels?tab=overview) as model data on a 0.11° resolution. As reference we use [ERA5-land](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land?tab=overview) data remapped to Cordex resolution using conservative remapping. Both Cordex and ERA5-land are publicy available. The data is available in the package as `bc_data` and includes daily reference and model data for a calibration period of 1990-2009 (`rc`and `mc`), as well as for a projection period of 2010-2019 (`rp`and `mp`). In addition, coordinates of the 22 grid points are provided. We will bias correct the data using the `gn_vbc` function.


The example code demonstrates how to apply the `GN-VBC` method to this dataset, including fitting GAMs, extracting components, fitting nested vine copulas, and performing bias correction.

```{r lib_n_data, message=FALSE}
library(data.table)
library(ggplot2)
library(knitr)
library(VBC)
library(dplyr)
library(tidyr)
library(patchwork)
library(GNVBC)

data("bc_data")

names_bc = names(bc_data)

vars <- c("tas", "hurs", "sfcWind", "ps")

bc_data[1:4] <- lapply(bc_data[1:4], function(df) {
  df %>%
    filter(between(time, as.Date("2000-01-01"), as.Date("2015-12-31")))
})

names(bc_data) = names_bc

```

In this example, we cut the timeframes and keep three variables only. To give an idea about the distribution of the data, we show differences in mean temperature between reference and model projections. Note that since the time periods are chosen differently in the main paper and another reference is used (here ERA5-land, data provided by MeteoSwiss in the paper), the figure shows slightly different values.


```{r description, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.align='center'}
reshape_dataset <- function(df, name, var = "tas") {
  
  df %>%
    select(time, starts_with(paste0(var, "."))) %>%
    pivot_longer(
      -c(time),
      names_to = "Id",
      names_prefix = paste0(var, "."),
      values_to = "value"
    ) %>%
    mutate(Id = as.integer(Id)) %>%
    group_by(Id) %>%
    summarise(mean = mean(value, na.rm = TRUE), .groups = "drop") %>%
    mutate(dataset = name) %>%
    left_join(bc_data$locations, by = "Id")
}
plot_df <- purrr::imap_dfr(bc_data[c("rc", "rp", "mc", "mp")], reshape_dataset)

# ---- Add Type + Scenario ----
plot_df <- plot_df %>%
  mutate(
    Type = ifelse(dataset %in% c("rc", "rp"), "Reference", "Model"),
    Scenario = ifelse(grepl("c$", dataset), "Calibration", "Projection")
  )

# ---- Compute Differences ----
df_plot <- plot_df %>%
  select(Id, Lon, Lat, Type, Scenario, mean) %>%
  pivot_wider(names_from = Type, values_from = mean) %>%
  mutate(`Model - Reference` = Model - Reference) %>%
  pivot_longer(
    cols = c(Reference, `Model - Reference`),
    names_to = "RowGroup",
    values_to = "value"
  )

# ---- Load Vaud Shape ----
swiss_cantons <- rnaturalearth::ne_states	(country = "Switzerland", returnclass = "sf")
vaud <- swiss_cantons %>% filter(name == "Vaud")

# ---- Plot ----
p_ref <- ggplot() +
  geom_sf(data = vaud, fill = "lightgrey", color = "black", linewidth = 1.2) +
  geom_point(
    data = df_plot %>% filter(RowGroup == "Reference"),
    aes(Lon, Lat, color = value),
    size = 6
  ) +
  facet_wrap(~Scenario, nrow = 1) +
  coord_sf(xlim = c(5.9, 7.4), ylim = c(46.1, 47.1)) +
  scale_color_viridis_c(
    option = "rocket",
    direction = -1,
    name = "Mean Temperature (K)"
  ) +
  xlab("Reference") + ylab("") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0)
  )

# -----------------------------
# Difference plot
# -----------------------------
p_diff <- ggplot() +
  geom_sf(data = vaud, fill = "lightgrey", color = "black", linewidth = 1.2) +
  geom_point(
    data = df_plot %>% filter(RowGroup == "Model - Reference"),
    aes(Lon, Lat, color = value),
    size = 6
  ) +
  facet_wrap(~Scenario, nrow = 1) +
  coord_sf(xlim = c(5.9, 7.4), ylim = c(46.1, 47.1)) +
  scale_color_gradient2(
    low = "slateblue2",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-4, 3.1),
    name = "Difference (K)"
  ) +
  xlab("Model − Reference") + ylab("") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0)
  )

# -----------------------------
# Combine
# -----------------------------
p_ref + p_diff
```
We observe that the model underestimates mean temperatures in the Eastern part of the canton, while it overestimates them in the Western part, clearly giving an incentive for bias correction. The differences are similar for both the calibration and projection period.

In order to assess distributional similarity, we compute Wasserstein distances between the model and reference data for each variable. This serves as an indicator for preservation of spatial dependence.

```{r wd_pre, message=FALSE}
wd_pre <- list()

# Loop over variables and calculate spatial Wasserstein distances
for (var in vars) {
  wd <- VBC::calc_wasserstein(
    bc_data$rp %>% select(starts_with(paste0(var, "."))),
    bc_data$mp %>% select(starts_with(paste0(var, ".")))
  )
  
  wd_pre[[var]] <- wd
}

wd_df <- do.call(cbind, wd_pre) %>% as.data.frame()
colnames(wd_df) <- vars
rownames(wd_df) <- c("WD 1", "WD 2")  


knitr::kable(wd_df, digits = 3, caption = "Wasserstein distances prior to bias correction")
```

In the next step, we apply the `gn_vbc` function to bias correct the model data. Note that this is a computationally intensive step, especially for larger datasets, and may take some time to run.

```{r mbc, echo = FALSE}
mp_corrected = gn_vbc(
  mp = bc_data$mp %>% select(-time),
  mc = bc_data$mc %>% select(-time),
  rc = bc_data$rc %>% select(-time),
  time_c = bc_data$mc$time,
  time_p = bc_data$mp$time,
  locs = bc_data$locations,
  nrows = 7,
  ncols = 7,
  families = list(
    "tas" = gaussian(),
    "hurs" = betar(),
    "sfcWind" = Gamma(link = "log"),
    "ps" = gaussian()
  ),
  fixed = FALSE
)
```


We calculate Wasserstein distances again to assess the improvement in distributional similarity after bias correction.
```{r wd_post, echo = FALSE}
wd_post <- list()
# Loop over variables and calculate spatial Wasserstein distances
for (var in vars) {
  wd <- VBC::calc_wasserstein(
    bc_data$rp %>% select(starts_with(paste0(var, "."))),
    mp_corrected$corrected_mp %>% select(starts_with(paste0(var, ".")))
  )
  
  wd_post[[var]] <- wd
}

wd_df <- do.call(cbind, wd_post) %>% as.data.frame()
colnames(wd_df) <- vars
rownames(wd_df) <- c("WD 1", "WD 2")  

```

```{r wd_improv}
# Calculate improvement in Wasserstein distance
wd_improvement <- sapply(vars, function(var) {
  pre <- wd_pre[[var]]
  post <- wd_post[[var]]
  improvement <- (pre - post)/pre * 100
  return(improvement)
})

knitr::kable(wd_improvement, digits = 1, caption = "Improvement in Wasserstein distances (%)")

```
We can visualize the resulting first tree of the estimated copulas for both `mp` (middle) and `rc` (bottom).
```{r visualize_trees, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.align='center'}
library(rvinecopulib)

# Switzerland cantons
swiss_cantons <- rnaturalearth::ne_states(country = "Switzerland", returnclass = "sf")
vaud <- swiss_cantons[swiss_cantons$name == "Vaud", ]

points_sf <- sf::st_as_sf(
  bc_data$locations,
  coords = c("Lon", "Lat"),
  crs = 4326
)

ggplot() +
  geom_sf(data = vaud, fill = "grey90", color = "black", linewidth = 1) +
  geom_sf(data = points_sf, color = "blue", size = 3) +
  geom_text(
    data = bc_data$locations,
    aes(x = Lon, y = Lat, label = Id),
    vjust = -1,
    size = 4
  ) +
  coord_sf(
    xlim = c(5.9, 7.4),
    ylim = c(46.15, 47.05)
  ) +
  theme_bw() +
  labs(title = "",
       x = "",
       y = "")

p_mp = plot(mp_corrected$rvine_mp, var_names = "use", main = "mp")
p_rc = plot(mp_corrected$rvine_rc, var_names = "use", main = "rc")



```

Finally, we can visualize the bias-corrected mean temperature values per grid cell and compare them to the reference data.

```{r mean_maps, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.align='center'}
plot_df <- purrr::imap_dfr(list("rp" = bc_data$rp, "mp" = cbind("time" = bc_data$rp$time, mp_corrected$corrected_mp)), reshape_dataset)

# ---- Add Type + Scenario ----
plot_df <- plot_df %>%
  mutate(
    Type = ifelse(dataset %in% c("rc", "rp"), "Reference", "Model"),
    Scenario = ifelse(grepl("c$", dataset), "Calibration", "Projection")
  )

# ---- Compute Differences ----
df_plot <- plot_df %>%
  select(Id, Lon, Lat, Type, Scenario, mean) %>%
  pivot_wider(names_from = Type, values_from = mean) %>%
  mutate(`Model - Reference` = Model - Reference) %>%
  pivot_longer(
    cols = c(Reference, `Model - Reference`),
    names_to = "RowGroup",
    values_to = "value"
  )

# ---- Load Vaud Shape ----
swiss_cantons <- rnaturalearth::ne_states	(country = "Switzerland", returnclass = "sf")
vaud <- swiss_cantons %>% filter(name == "Vaud")

# ---- Plot ----
p_ref <- ggplot() +
  geom_sf(data = vaud, fill = "lightgrey", color = "black", linewidth = 1.2) +
  geom_point(
    data = df_plot %>% filter(RowGroup == "Reference"),
    aes(Lon, Lat, color = value),
    size = 6
  ) +
  facet_wrap(~Scenario, nrow = 1) +
  coord_sf(xlim = c(5.9, 7.4), ylim = c(46.1, 47.1)) +
  scale_color_viridis_c(
    option = "rocket",
    direction = -1,
    name = "Mean Temperature (K)"
  ) +
  xlab("Reference") + ylab("") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0)
  )

# -----------------------------
# Difference plot
# -----------------------------
p_diff <- ggplot() +
  geom_sf(data = vaud, fill = "lightgrey", color = "black", linewidth = 1.2) +
  geom_point(
    data = df_plot %>% filter(RowGroup == "Model - Reference"),
    aes(Lon, Lat, color = value),
    size = 6
  ) +
  facet_wrap(~Scenario, nrow = 1) +
  coord_sf(xlim = c(5.9, 7.4), ylim = c(46.1, 47.1)) +
  scale_color_gradient2(
    low = "slateblue2",
    mid = "white",
    high = "red3",
    midpoint = 0,
    limits = c(-4, 3.1),
    name = "Difference (K)"
  ) +
  xlab("Model − Reference") + ylab("") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0)
  )

p_diff_rescaled <- ggplot() +
  geom_sf(data = vaud, fill = "lightgrey", color = "black", linewidth = 1.2) +
  geom_point(
    data = df_plot %>% filter(RowGroup == "Model - Reference"),
    aes(Lon, Lat, color = value),
    size = 6
  ) +
  facet_wrap(~Scenario, nrow = 1) +
  coord_sf(xlim = c(5.9, 7.4), ylim = c(46.1, 47.1)) +
  scale_color_gradient2(
    low = "slateblue2",
    mid = "white",
    high = "red3",
    midpoint = 0,
    # limits = c(-4, 3.1),
    name = "Difference (K)"
  ) +
  xlab("Model − Reference") + ylab("") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0)
  )
# -----------------------------
# Combine
# -----------------------------
p_ref + p_diff + p_diff_rescaled
```
We see that the bias correction has successfully reduced the differences in mean temperature between model and reference. In the middle panel, we apply the same range of differences as in the plot above, while in the right panel we zoom in to detect small deviations from the reference. The improvement in Wasserstein distances indicates that the bias correction has also improved the distributional similarity between model and reference data, suggesting that spatial dependence structures have been better preserved.

## Citation
If you use `GN-VBC` in a scientific publication, please cite the following paper:

[add archive link]

## References
Funk, H., Ludwig, R., Küchenhoff, H., Nagler, T. (2025). Towards more realistic climate model outputs: a multivariate bias correction based on zero-inflated vine copulas. Journal of the Royal Statistical Society Series C: Applied Statistics; qlaf044.

Copernicus Climate Change Service (C3S)(2019): ERA5-Land hourly data from 1950 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). DOI: 10.24381/cds.e2161bac (Accessed on 19-03-2025).

Copernicus Climate Change Service, Climate Data Store, (2019): CORDEX regional climate model data on single levels. Copernicus Climate Change Service (C3S) Climate Data Store (CDS). DOI: 10.24381/cds.bc91edc3 (Accessed on 13-02-2025).

